Transfer learning for echelle spectroscopy

### 2021-11-10 paper/paper1/ms.tex
writing prose

We start with a high resolution pre-computed stellar model spectrum, $\mathsf{M}(\bm{\lambda})$.  This spectrum should have sporadic regions of discernable continuum devoid of lines, and the continuum must vary smoothly in wavelength.  We suspect most stellar spectra meet these criteria, with the exception of the coolest M dwarfs and brown dwarfs.  The method can hypothetically handle spectra without discernable continua, but we suspect some modifications to the preprocessing steps would be needed.  We allow the spectrum to have any number of spectral lines.  We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$, chosen to account for plausible radial velocity and rotational broadening of real stars.  A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

@@

We initialize the cloned model with a series of preprocessing steps.  We divide the entire spectrum by a black body of the same effective temperature $T_{\mathrm{eff}}$ as the model template.  The resulting signal typically possesses a smooth continuum variation, since

Occasionally a more advanced continuum flattening step may be needed to ensure that subsequent spectral line finding steps get applied uniformly.  Here, we apply a low-pass Savitzky-Golay filter\authorcomment1{citation XX} to identify the continuum regions, and then fit a low-order polynomial to this pre-filtered continuum spectrum.  We evaluate the low-order polynomial at all wavelength points, multiplied by the blackbody curve to assemble a ``perturbed black body'' that symbolizes the black body modulated by any broad-band continuum opacity that may reside in the model.  We define a threshold flux loss of $1\%$, below which any lines are masked, erode that mask by $N_{\mathrm{iter}}=13$ pixels, and repeat the polynomial fitting process again.  At this stage we fit a high order polynomial, with the polynomial order set through cross validation, typically $n_{\mathrm{poly}}\sim11$.  The ``perturbed black body'' continuum model $\mathsf{B}$ contains  $n_{\mathrm{poly}}+1$ free parameters.

In either continuum flattening case, we move on to find locations of spectral lines.
We apply a local minimum finding algorithm on the spectrum by defining a prominence threshold $\delta f\sim0.01$.
Testing demo with the real fixed static residual.

>>>>

We start with a high resolution pre-computed stellar model spectrum, $\mathsf{M}(\bm{\lambda})$.  The procedure is largely agnostic to the exact details of how this spectrum was made.  For the purposes of this paper, we will showcase examples from two well-known families of precomputed synthetic spectra: \texttt{PHOENIX} \citep{husser13} for stellar spectra $(T_{\mathrm{eff}}\in [2300, 10000]\;K)$ and \texttt{Sonora} \citep{2021ApJ...920...85M} for brown dwarf spectra $(T_{\mathrm{eff}}\in [1000, 2300]\;K)$.  We place the following demands on the the precomputed spectra.  They should have sporadic regions of discernable continuum devoid of lines, and the continuum must vary smoothly in wavelength.  The spectral lines or pseudo-lines should be resolved, and not sub-sampled.  We suspect most stellar spectra meet these criteria, with the exception of the coolest M dwarfs and brown dwarfs.  The method can hypothetically handle spectra without discernable continua, but we suspect some modifications to the preprocessing steps would be needed.  We hypothetically allow the spectrum to have any number of spectral lines, though in practice there are computational tradeoffs as the number of lines grows.  We truncate the red and blue limits of the precomputed synthetic spectrum to match a high-bandwidth echelle spectrograph, extended with a buffer at the edges of size $\pm \Delta \lambda_{\mathrm{buffer}}$, chosen to account for plausible radial velocity and rotational broadening of real stars.  A generous buffer of $v \sin{i} < 500 \;\kmps$ and $|RV|<500 \;\kmps$ yields a typical buffer of about 30~\AA.

@@

We initialize the cloned model with a series of preprocessing steps.  We divide the entire spectrum by a black body of the same effective temperature $T_{\mathrm{eff}}$ as the model template.  The resulting signal typically possesses a smooth continuum variation.  An additional continuum flattening step ensures that subsequent spectral line finding steps get applied uniformly.  Here, we identify the continuum regions as the most prominent peaks in the specturm separated by a pixel distance of at least $15000-90000$ pixels from adjacent peaks, yielding between 60 and 5 points.  We then fit an $n_{\mathrm{poly}} = 3^{rd}-5^{th}$ order polynomial to these peaks, evaluate the polynomial on the entire bandwidth, and divide the spectrum by this trend.  Visual inspection of this continuum-flattened spectrum typically possesses variations at the $<0.2\%$ level.  Tuning the pixel distance and polynomial order can lower the undulations to $<0.05\%$ level for some input spectra.

We emphasize that a recreation of the unvarnished input spectrum can be obtained by multiplying by the blackbody curve times this polynomial shape, yielding the ``perturbed black body'', $\mathsf{B}(\bm{\lambda})$, that symbolizes the black body modulated by continuum opacity or broad-band radiative transfer effects.  This smooth spectrum may be useful for applications that need to keep track of broad-band flux, such as low resolution, or molecular bandheads.  The ``perturbed black body'' continuum model $\mathsf{B}$ contains  $n_{\mathrm{poly}}+1$ fixed lookup parameters, plus the input $T_{\mathrm{eff}}$.

We apply a local-minimum-finding algorithm on the spectrum by defining a prominence threshold $P_{rom} \in (0.005, 0.1)$.  This threshold dictates the number of lines that will be modeled: a lower prominence finding more, smaller lines, and a larger prominence finding fewer, deeper lines.  The prominence algorithm successfully finds lines that reside on top of broad line wings, provided that they exceed the prominence threshold in their local region.  Figure XX shows how
